Multi-scale assessment of a grassland productivity model

### 2020-12-28 manuscript/manuscript.Rmd
minor edits/typos

A promising method for this is low dimensional models, which are process models with some simplified components [@choler2010; @choler2011]. For example, a low dimensional model might approximate transpiration to a function of potential evapotranspiration, soil available water, and live vegetation cover along with a single parameter. As opposed to a high dimensional model with multiple functions accounting for leaf area index, stomatal conductance, rooting depth and surface area, etc [@caylor2009; @asbjornsen2011]. The low dimensional model is advantageous since it can generalize across broad regions with relatively few inputs. Yet they are still susceptible to over-fitting to local conditions since parameters or model structure can be tied to specific locations or plant functional groups [@fisher2020]. Thus parameterizing low-dimensional models must be done with care such that they are applicable to a broad area while maintaining an acceptable level of accuracy.

@@

Models which exceeded the threshold were subject to further evaluation. For each model we performed a leave one out cross-validation, where the model was re-fit with one Phenocam site not included in the training data, and then evaluated against this left out site. In this step a scaling coefficient to link mean annual precipitation with PhenoCam G\textsubscript{cc} was held constant at the value obtained in first fitting. The resulting \(\overline{NSE}\) and \textit{F} are from all modelled sites using their respective out of sample test.

@@

At the largest spatial scale, where the PhenoGrass model was parameterized using all 89 sites, the model performed poorly with an \(\overline{NSE}\) value among sites of 0.31 (Table 1; Fig. 3). Models built using all sites of a respective vegetation type performed poorly as well, though were slightly better than the all site model (Fig. 3). The best model performance was achieved when models were built using a single vegetation type subset to a single ecoregion. Grasslands within the Great Plains and Eastern Temperate Forests ecoregions were the only instances where \(\overline{NSE}\) exceeded the 0.65 threshold, though Grasslands within N.W. Forests came close (NSE = 0.64).

@@

The grassland vegetation type, subset to specific ecoregions, predominantly outperformed models built with other spatial scales (Table 1). Models built using grasslands within the Eastern Temperate Forest and Great Plains ecoregions had the highest \(\overline{NSE}\) values of 0.82 and 0.69, respectively. Using leave one out cross-validation on these two grassland models resulted in similar errors of 0.79 and 0.67 for the Eastern Temperate Forest and Great Plains, respectively. Though N.W. Forests grasslands had an in-sample \(\overline{NSE}\) just below the 0.65 threshold, the cross-validation was well below it (0.52). Grasslands in the North American deserts were not modelled well at any spatial scale and had the lowest \(\overline{NSE}\) values in the entire analysis. The observed greenness patterns of these desert grasslands had extremely high variability in their magnitude and timing, with short distinct peaks in greenness and numerous off-peak fluctuations. The fitted model, which minimized the mean CVME among the 5 sites, was not able to reproduce this high variability and instead produced fCover values that were severely constrained to a narrow range (Fig. S1).

@@

We performed an extensive evaluation of the PhenoGrass model across ecoregions and vegetation types to determine the best spatial scale at which to parameterize and apply the model. We found the model most suitable to grassland vegetation when constrained to the ecoregion level, though it did not perform well in grasslands in the North American desert ecoregion. Shrublands and agriculture were not well represented by the model regardless of the spatial scale. Results from this study will facilitate long-term projections of grassland productivity constrained to an appropriate vegetation type and scale.

@@

@hufkens2016 originally evaluated the PhenoGrass model using 14 grassland sites distributed among seven North American ecoregions. In their evaluation they had an average R\textsuperscript{2} of 0.71, while here the model performed poorly when using more than 1 ecoregion. It’s likely that the original 14 grassland sites were ideal locations for the PhenoGrass model, since on average they have a single greenup season every year in the spring or summer (Fig. 5A). The additional 24 grassland sites used in the current study have high seasonal variability and elongated growing seasons (Fig. 5B, Fig. S1), and were thus more difficult to represent in a single continental scale grassland model. This highlights the need for longer time series in evaluating low-dimensional models as it may take many years for a single location to experience the full range of variability. As the PhenoCam data archive grows, temporally out of sample validation can be done to better evaluate performance into novel conditions.

>>>>

A promising method for this is low-dimensional models, which are process based models with some simplified components [@choler2010; @choler2011]. For example, a low-dimensional model might approximate transpiration as a function of potential evapotranspiration, soil available water, and live vegetation cover along with a single parameter. As opposed to a high-dimensional model with multiple functions accounting for leaf area index, stomatal conductance, rooting depth and surface area, etc [@caylor2009; @asbjornsen2011]. The low-dimensional model is advantageous since it can generalize across broad regions with relatively few inputs. Yet they are still susceptible to over-fitting to local conditions since parameters or model structure can be tied to specific locations or plant functional groups [@fisher2020]. Thus parameterizing low-dimensional models must be done with care such that they are applicable to a broad area while maintaining an acceptable level of accuracy.

@@

Models which exceeded the threshold were subject to further evaluation. For each model we performed a leave one out cross-validation, where the model was re-fit with one Phenocam site not included in the training data, and then evaluated against this left out site. In this step a scaling coefficient to link mean annual precipitation with PhenoCam G\textsubscript{cc} was held constant at the value obtained in the first fitting. The resulting \(\overline{NSE}\) and \textit{F} are from all modelled sites using their respective out of sample test.

@@

At the largest spatial scale, where the PhenoGrass model was parameterized using all 89 sites, the model performed poorly with an \(\overline{NSE}\) value of 0.31 (Table 1; Fig. 3). Models built using all sites of a respective vegetation type performed poorly as well, though were slightly better than the all site model (Fig. 3). The best model performance was achieved when models were built using a single vegetation type subset to a single ecoregion. Grasslands within the Great Plains and Eastern Temperate Forests ecoregions were the only instances where \(\overline{NSE}\) exceeded the 0.65 threshold, though Grasslands within N.W. Forests came close (NSE = 0.64).

@@

The grassland vegetation type, subset to specific ecoregions, predominantly outperformed models built with other spatial scales (Table 1). Models built using grasslands within the Eastern Temperate Forest and Great Plains ecoregions had the highest \(\overline{NSE}\) values of 0.82 and 0.69, respectively. Using leave one out cross-validation on these two grassland models resulted in similar errors of 0.79 and 0.67 for the Eastern Temperate Forest and Great Plains, respectively. Though N.W. Forests grasslands had an in-sample \(\overline{NSE}\) just below the 0.65 threshold, the cross-validation was well below it (0.52). Grasslands in the North American deserts were not modelled well at any spatial scale and had the lowest \(\overline{NSE}\) values in the entire analysis. The observed greenness patterns of these desert grasslands had extremely high variability in their magnitude and timing, with short distinct peaks in greenness and numerous off-peak fluctuations. The fitted model, which minimized \textit{F} among the 5 sites, was not able to reproduce this high variability and instead produced fCover values that were severely constrained to a narrow range (Fig. S1).

@@

We performed an extensive evaluation of the PhenoGrass model across ecoregions and vegetation types to determine the best spatial scale at which to parameterize and apply the model. We found the model most suitable to grassland vegetation when constrained to the ecoregion level, though it did not perform well in grasslands in the North American Deserts ecoregion. Shrublands and agriculture were not well represented by the model regardless of the spatial scale. Results from this study will facilitate long-term projections of grassland productivity constrained to an appropriate vegetation type and scale.

@@

@hufkens2016 originally evaluated the PhenoGrass model using 14 grassland sites distributed among seven North American ecoregions. In their evaluation they had an average R\textsuperscript{2} of 0.71, while here the model performed poorly when using more than one ecoregion. It’s likely that the original 14 grassland sites were ideal locations for the PhenoGrass model, since on average they have a single greenup season every year in the spring or summer (Fig. 5A). The additional 24 grassland sites used in the current study have high seasonal variability and elongated growing seasons (Fig. 5B, Fig. S1), and were thus more difficult to represent in a single continental scale grassland model. This highlights the need for longer time series in evaluating low-dimensional models as it may take many years for a single location to experience the full range of variability. As the PhenoCam data archive grows, temporally out of sample validation can be done to better evaluate performance into novel conditions.
